### 2022-09-06 latex/Chapters/DRLSearch.tex
update

Since my implementation of \textbf{DQL} is no longer just a cost-to-go estimate as with \textbf{DRL}, but a joint cost-to-go and distribution over actions (to be concrete, that means a 5-dimensional vector for the \textbf{SP} and a 13-dimensional vector for the \textbf{RC}). This leaves us with the question of how can we use the output of such a network to search for a solution. One simple answer (which I have tried, see e.g. subsection \ref{sec:S33DRLDQL}) is that we could disregard the distribution and only take into account the cost-to-go by simply using $A^{*}$. The rationale for doing so is that we might hope that jointly learning the cost-to-go and the actions via a combination of MSE and cross-entropy-loss produces a better network and leads to better heuristics, since we use more information during the learning and iterations.
\\
\\
Another possibility is to use an algorithm which combines both the actions distribution and the value function to inform its search. The Monte Carlo Tree Search (\textbf{MCTS}) is such an algorithm. In short, it generates at each step (and possibly in a distributed manner) paths to a \textit{leaf} on the frontier of unexpanded nodes, expands that node and checks if it is a goal state (and keeps going until the goal is found). The paths followed from initial state to leaves are decided via a combination of the probability distribution over actions (following the most promising actions) as well as the value function, according to a trade-off which can be tuned via a hyper-parameter. In order to prevent always following the same paths (which might take us very far from the goal), \textbf{MCTS} algorithms usually keep track of actions they have followed already and underweight their probability so that subsequent paths explore other actions. A pseudo code description of it looks like this:
\\
\teal
\paragraph{}{\textbf{pseudo code -- \textbf{MCTS}}}
\begin{pseudocode}
############################################################
def MCTS(initial_node, time_out, q_v_network):
    """ pseudo code for Monte Carlo Tree Search """
    initialize_time_out(time_out)
    tree = Tree(initial_node)
    while True:
        check_time_out()
        leaf = add_new_leaf(tree, q_v_network)
        if leaf.is_goal():
            return BFS(tree) # path to leaf can usually be improved by BFS
############################################################
def add_new_leaf(tree, q_v_network):
    leaf = tree.initial_node
    while leaf.expanded:
        pass
    leaf.expand() # add children of leaf to tree
    leaf.update_penalties()
    return leaf
############################################################
\end{pseudocode}
\black

I have implemented the exact \textbf{MCTS} procedure described in \cite{https://doi.org/10.48550/arxiv.1805.07470} and outlined in pseudo code above in a class called MonteCarloTreeSearch (see section \ref{sec:codesolvers} for details).

>>>>

Since my implementation of \textbf{DQL} is no longer just a cost-to-go estimate as with \textbf{DRL}, but a joint cost-to-go and distribution over actions (to be concrete, that means a 5-dimensional vector for the \textbf{SP} and a 13-dimensional vector for the \textbf{RC}). This leaves us with the question of how can we use the output of such a network to search for a solution. One simple answer (which I have tried, see e.g. subsection \ref{sec:S33DRLDQL}) is that we could disregard the distribution and only take into account the cost-to-go by simply using $A^{*}$. The rationale for doing so is that we might hope that jointly learning the cost-to-go and the actions via a combination of MSE and cross-entropy-loss produces a better network and leads to better heuristics, since we use more information during the learning and iterations.
\\
\\
Another possibility is to use an algorithm which combines both the actions distribution and the value function to inform its search. The Monte Carlo Tree Search (\textbf{MCTS}) is such an algorithm. In short, it generates at each step (and possibly in a distributed manner) paths to a \textit{leaf} on the frontier of unexpanded nodes, expands that node and checks if it is a goal state (and keeps going until the goal is found). The paths followed from initial state to leaves are decided via a combination of the probability distribution over actions (following the most promising actions) as well as the value function, according to a trade-off which can be tuned via a hyper-parameter. In order to prevent always following the same paths (which might take us very far from the goal), \textbf{MCTS} algorithms usually keep track of actions they have followed already and underweight their probability so that subsequent paths explore other actions. A pseudo code description of it looks like this:
\\
\teal
\paragraph{}{\textbf{pseudo code -- \textbf{MCTS}}}
\begin{pseudocode}
#################################################################
def MCTS(initial_node, time_out, q_v_network):
    """ pseudo code for Monte Carlo Tree Search """
    initialize_time_out(time_out)
    tree = Tree(initial_node)
    while True:
        check_time_out()
        leaf = construct_path_to_new_leaf(tree, q_v_network)
        if leaf.is_goal():
            return BFS(tree) # path to leaf can usually be improved by BFS
#################################################################
def construct_path_to_new_leaf(tree, q_v_network):
    leaf = tree.initial_node
    while leaf.expanded:
        actions_proba, value = q_v_network(leaf)
        historical_actions_count = leaf.actions_taken
        leaf = leaf.choose_child(actions_proba,
                                 value,
                                 historical_actions_count)
    # add children of leaf to tree
    leaf.expand()
    # penalize path taken to favour exploration
    leaf.increment_historical_actions_count()
    return leaf
#################################################################
\end{pseudocode}
\black

I have implemented the exact \textbf{MCTS} procedure described in \cite{https://doi.org/10.48550/arxiv.1805.07470} and outlined in the above pseudo code in  a class called MonteCarloTreeSearch (see section \ref{sec:codesolvers} for details).

### 2022-09-06 latex/Chapters/Puzzles.tex
update

%-----------------------------------
%	SECTION 3
%-----------------------------------
\section{Solving puzzles \textit{without human knowledge}}

In the upcoming implementation (\ref{sec:Implementation}) and results (\ref{sec:ResultsSP}, \ref{sec:ResRubiks}) sections, I will be discussing various solvers. Some of them, such as my implementation of a naive \textbf{SP} solver, or the Kociemba open source implementation for the \textbf{RC}, make heavy use of specific knowledge about these two puzzles. Then there are the A$^{*}$ with handcrafted heuristics (e.g. Manhattan heuristic for the \textbf{SP}) algorithms which .

, which I have taken from third parties (kociemba) to supervised (DL) to unsupervised (DRL \& DQL) .... but sometimes subtle knowledge injected in e.g. representation or simplifications.

>>>>

%-----------------------------------
%	SECTION 3
%-----------------------------------
\section{Solving puzzles without human knowledge}
\label{sec:WHK}

In the upcoming implementation (\ref{sec:Implementation}) and results (\ref{sec:ResultsSP}, \ref{sec:ResRubiks}) sections, I will be discussing various solvers. I would put them broadly in four different categories, by decreasing level of human knowledge that has been assumed and handcrafted in them. Notice that this human-knowledge-required scale is obviously somewhat arbitrary, but good to keep in mind in later (and in particular results) sections.

\paragraph{Domain-specific solvers}
At 4 on the human-knowledge-required scale, I would put my NaiveSolver (handcrafted to solve the \textbf{SP}) as well as the hkociemba (2x2x2) and kociemba (3x3x3) open source implementations which I have integrated in my KociembaSolver. They make heavy use of specific knowledge about the two puzzles, not only about their mechanics, but also about how to actually (provably) solve them.

\paragraph{Domain-specific heuristics}
At 3 on the human-knowledge-required scale, I put A$^{*}$-with-handcrafted-heuristics (e.g. Manhattan heuristic for the \textbf{SP}) algorithms. Constructing the heuristics usually require some knowledge about the puzzle's mechanics, but not necessarily about how to solve them efficiently (though of course knowing the latter might help come up with more efficient heuristics).

\paragraph{Supervised learning}
At 2 on the human-knowledge-required scale, I put my DeepLearner solver (also using A$^{*}$ but whose heuristic is learnt via \textbf{DL}). This solver does not require any knowledge of the structure or mechanics of the puzzle, nor of how to solve them, but just need a teacher algorithm to learn/extrapolate from.

\paragraph{Blind search\& unsupervised learning}
Finally at 1 on the human-knowledge-required scale, I put the blind solvers such as \textbf{BFS} and \textbf{DFS}. Also in that group I put A$^{*}$ with heuristics learnt using my DeepReinforcementLearner and DeepQLearner, and my MonteCarloSearchTree which all know (almost) nothing about the puzzles and do totally work from generic interfaces alone (states, transitions, etc \dots) and are operating in an unsupervised way.
\\
\\
One important comment I would make to end this section is that it is easy to inject, without realising or acknowledging it, specific knowledge into the solvers, or in the tensor representation of the puzzles that the solvers take as input. Pretty much all the papers about \textbf{RC} that I have surveyed chose a more compact representation for the cube than I have. They remove the center cubies of the 3x3x3 cube since, they argue, those never move via face-turns. Some of them go much further and make use of various symmetry arguments to reduce the search space and reduce the \textbf{RC} to having one goal state (instead of 24). One of the main papers I have inspired myself for this project, \cite{DBLP:journals/corr/abs-1805-07470}, reduces the representation of the Rubik's considerably more, by making use of arguments about the physical structure of the edge and corner cubies: knowing the color of one of their face allows you do deduct the other faces, and therefore, they only need to track one face. I of course understand that they do that for the sake of reducing the size of the tensors representing the cube, and in order to remove redundant information. This is still \textit{cheating} a bit in my opinion, especially if you are claiming to solve these puzzles \textit{without-human-knowledge}, as we can only operate these dimensionality reduction tricks if we have external world knowledge about the physical structure of the puzzles. In this project, I have chosen as pure an approach as I could, and not only did not reduce the dimensionality of the tensor representation of the puzzles, but also, specifically for the Rubik's cube, I did not make use of invariance by spacial rotation of the full cube.

### 2022-09-06 latex/MScAIFrancoisBerrier.tex
update

\newcommand{\red}{\color{red}}
\newcommand{\black}{\color{black}}
\newcommand{\blue}{\color{blue}}
\definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
\newcommand{\afblue}{\color{airforceblue}}

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
\textbf{Motivation}
\label{Abstract}
\\
Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in the Sutton book (\cite{Sutton1998}), has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods.
\\
Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubik's Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQL})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using (\textbf{DL}) or (\textbf{DRL}), to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis}
\\
In the first chapter, I will quickly describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In chapter 2, I will do a quick recap of the different methods that I will use in the project. Chapter 3 will be dedicated to discussing the mathematics of the sliding puzzle (\textbf{SP}) and the (\textbf{RC}). I might throw a few random (but hopefully interesting) observations in there and give some references for the keen reader. I will then give an overview in chapter 4 of the code base I have developed - and put in the open on my github page (\cite{FB}) - to complete this project before detailing a few examples in chapter 5. Finally, in chapters 6 and 7, I will present all my various results on respectively the sliding puzzle and the Rubik's cube before offering some conclusion in chapter 8.

>>>>

\newcommand{\red}{\color{red}}
\newcommand{\black}{\color{black}}
\newcommand{\blue}{\color{blue}}
\definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
\newcommand{\afblue}{\color{airforceblue}}
\newcommand{\teal}{\color{teal}}

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
\textbf{Motivation}
\label{Abstract}
\\
Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in the Sutton book (\cite{Sutton1998}), has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods.
\\
Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubik's Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQL})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using (\textbf{DL}) or (\textbf{DRL}), to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis}
\\
In the first chapter, I will quickly describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In chapter 2, I will do a quick recap of the different methods that I will use in the project. Chapter 3 will be dedicated to discussing the mathematics of the sliding puzzle (\textbf{SP}) and the (\textbf{RC}). I might throw a few random (but hopefully interesting) observations in there and give some references for the keen reader. I will then give an overview in chapter 4 of the code base I have developed - and put in the open on my github page (\cite{FB}) - to complete this project. Finally, in chapters 6 and 7, I will present all my various results on respectively the sliding puzzle and the Rubik's cube before offering some conclusion in chapter 8.
\\
A first appendix details how to use the code base, from constructing puzzles, to learning and solving them using all the different methods I have implemented for this project. A second appendix describes a limitation I bumped into while using the kociemba library (an open source 3x3x3 \textbf{RC} solver), and how I circumvented it.
